---
title: "Drosophila melanogaster pupae TFL: summary of progress and future plans"
author: "Benjamin Walsh"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    fig_caption: yes
    toc: true
    toc_float: true
    theme: spacelab
    highlights: zenburn
---


```{r setup, include=FALSE}
#### Loading packages and data in ####
rm(list = ls()) #clear R

# Load packages and data ####
library(ggplot2)
library('Cairo')
library(dplyr)
library(binom)
library(tidyr)
library(here)

#Load data
setwd("M:\\Documents\\D_mel_pupae_TFL\\Data")
pupaeTFL<-read.csv("D.mel_pupae_TFL_DATA.csv", header=TRUE)


# EXPERIMENT 1: PUPAE TFL ####

#### Rename vectors and convert from numeric to factors #####
#pupaeTFL$temp<-as.factor(pupaeTFL$temp)
pupaeTFL$block<-as.factor(pupaeTFL$block)
pupaeTFL$PCR<-as.factor(pupaeTFL$PCR)
pupaeTFL$plate_ID<-as.factor(pupaeTFL$plate_ID)
pupaeTFL$vial_ID<-as.factor(pupaeTFL$vial_ID)
pupaeTFL$emerged<-as.factor(pupaeTFL$emerged)
pupaeTFL$sex<-as.factor(pupaeTFL$sex)
pupaeTFL$survived_to_mating<-as.factor(pupaeTFL$survived_to_mating)
pupaeTFL$M_male_survived_mating<-as.factor(pupaeTFL$M_male_survived_mating)
pupaeTFL$M_female_survived_mating<-as.factor(pupaeTFL$M_female_survived_mating)
pupaeTFL$M_female_survived_laying<-as.factor(pupaeTFL$M_female_survived_laying)
pupaeTFL$M_larval<-as.factor(pupaeTFL$M_larval)
#pupaeTFL$F_female_survived_mating<-as.factor(pupaeTFL$F_male_survived_mating)
pupaeTFL$F_male_survived_mating<-as.factor(pupaeTFL$F_male_survived_mating)
#pupaeTFL$F_female_survived_laying1<-as.factor(pupaeTFL$F_female_survived_laying1)
pupaeTFL$F_larval1<-as.factor(pupaeTFL$F_larval1)
#pupaeTFL$F_female_survived_laying2<-as.factor(pupaeTFL$F_female_survived_laying2)
pupaeTFL$F_larval2<-as.factor(pupaeTFL$F_larval2)
#pupaeTFL$F_female_survived_laying3<-as.factor(pupaeTFL$F_female_survived_laying3)
pupaeTFL$F_larval3<-as.factor(pupaeTFL$F_larval3)

# STATS ####
#### 1. Kruskal Wallis ####
#Kruskal Wallis with larval vs temp
#kruskal.test(M_larval ~ temp, data=pupaeTFL)

#### 2. Binomail GLM ####
#Basic GLM with larval vs temp in males
#M1<-glm(M_larval~temp,data=pupaeTFL, family='binomial')
#anova(M1,test="Chisq")
#summary(M1)

# DATA MANAGEMENT ####
#### 1. Creating subsets of data for plotting ####

#Subset of individuals with emergence data
subemerged<- subset(pupaeTFL,emerged==1 |emerged==0)
subemerged$emerged<-as.numeric(subemerged$emerged)
subemerged$emerged<-subemerged$emerged-1

#Create subset with male larval
subpupaeTFL<- subset(pupaeTFL,M_larval==1 |M_larval==0 |M_male_survived_mating ==1)
subpupaeTFL$M_larval<-as.numeric(subpupaeTFL$M_larval)
subpupaeTFL$M_larval<-subpupaeTFL$M_larval-1 #changing back to normal values

#Create subset with female larval where data for larval 3 is available
FsubpupaeTFL<- subset(pupaeTFL, F_larval3==1 |F_larval3==0)
FsubpupaeTFL$F_larval1<-as.numeric(FsubpupaeTFL$F_larval1)
FsubpupaeTFL$F_larval2<-as.numeric(FsubpupaeTFL$F_larval2)
FsubpupaeTFL$F_larval3<-as.numeric(FsubpupaeTFL$F_larval3)
FsubpupaeTFL$F_larval1<-FsubpupaeTFL$F_larval1-1
FsubpupaeTFL$F_larval2<-FsubpupaeTFL$F_larval2-1
FsubpupaeTFL$F_larval3<-FsubpupaeTFL$F_larval3-1

Fsublarval1<-subset(pupaeTFL, F_larval1==1 | F_larval1==0 & F_female_survived_laying1==1)
Fsublarval1$F_larval1<-as.numeric(Fsublarval1$F_larval1)
Fsublarval1$F_larval1<-Fsublarval1$F_larval1-1



#### 2. Calculating proportions for binary data ####
prop_emerged <- summarise(group_by(subemerged, temp), mean(emerged), N=length(emerged))
prop_M_larval <- summarise(group_by(subpupaeTFL, temp), mean(M_larval), N=length(M_larval))
prop_F_larvaltot <- summarise(group_by(FsubpupaeTFL, temp), mean(F_larval1)+mean(F_larval2)+mean(F_larval3), N=length(F_larval1))
prop_F_larval1<-summarise(group_by(Fsublarval1, temp), mean(F_larval1), N=length(F_larval1))

# PLOTS ####

#### 1. Emerged with error bars ####
prop_emerged<-cbind((binom.confint(prop_emerged$`mean(emerged)`*prop_emerged$N, prop_emerged$N, conf.level = 0.95, methods = "logit")),prop_emerged)

pd<-position_dodge(1)

emerged_error<- ggplot(prop_emerged, aes(x=temp, y=`mean(emerged)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=0.5, colour="black") +
  geom_line(position=pd, size=0.5, colour="black")+
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature (°C)', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of emerging individuals')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

emerged_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_emerged_error.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(emerged_error)
#dev.off() 

#### 2. Male fertility vs emergence error ####

prop_M_larval<-cbind((binom.confint(prop_M_larval$`mean(M_larval)`*prop_M_larval$N, prop_M_larval$N, conf.level = 0.95, methods = "logit")),prop_M_larval)

colnames(prop_M_larval)[8] <- "mean(emerged)"
prop_emerged_male <- rbind(prop_emerged,prop_M_larval)
prop_emerged_male [,"emerg_or_TFL"]<- c(0,0,0,0,0,0,0,0,1,1,1,1,1,1,1)
prop_emerged_male$emerg_or_TFL<-as.factor(prop_emerged_male$emerg_or_TFL)


pd0 <- position_dodge(0)
TFL_M_emergence_error <- ggplot(prop_emerged_male, aes(x=temp, y=`mean(emerged)`, group=emerg_or_TFL, ymin=lower, ymax=upper)) +
                         geom_point(position=pd0, size=1.5, aes(colour=factor(emerg_or_TFL))) +
                         geom_errorbar(position=pd0, size=0.5, width=.15, aes(colour=factor(emerg_or_TFL)))+
                         geom_line(position=pd0, aes(colour=factor(emerg_or_TFL)))+
                         scale_x_continuous('Pupal heat-shock temperature (°C)',breaks = round(seq(min(28), max(35), by = 1),1)) +
                         scale_y_continuous('Proportion')+
                         coord_cartesian(ylim=c(0, 1)) +
                         scale_color_discrete("",labels=c("Emergence", "Male fertility"))+
                         theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                               axis.line=element_line(colour='gray90'),
                               axis.text=element_text(colour='black'),
                               panel.grid.major = element_line("gray90"),
                               axis.title.y = element_text(vjust=1.5),
                               plot.title = element_text(face="bold"),
                               plot.background = element_rect(fill = "transparent",colour = NA))

TFL_M_emergence_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_TFL_M_emergence_error.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(TFL_M_emergence_error)
#dev.off() 

#### 3. Male fertility with error bars ####

#Calculate ymin and ymax


#Set position dodge
pd<-position_dodge(1)

TFL_M_error<- ggplot(prop_M_larval, aes(x=temp, y=`mean(emerged)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=0.5, colour="black") +
  geom_line(position=pd, size=0.5, colour="black")+
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature (°C)', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of surviving males fertile')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

TFL_M_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_TFL_M_error.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(TFL_M_error)
#dev.off() 

#### 4. Female fertility vs emergence error ####

prop_F_larval1<-as.data.frame(cbind((binom.confint(prop_F_larval1$`mean(F_larval1)`*prop_F_larval1$N, prop_F_larval1$N, conf.level = 0.95, methods = "logit")),prop_F_larval1))

colnames(prop_F_larval1)[8] <- "mean(emerged)"
prop_emerged_female <- rbind(prop_emerged,prop_F_larval1)
prop_emerged_female [,"emerg_or_TFL"]<- c(0,0,0,0,0,0,0,0,1,1,1,1,1,1,1)
prop_emerged_female$emerg_or_TFL<-as.factor(prop_emerged_female$emerg_or_TFL)


pd0 <- position_dodge(0)
TFL_F_emergence_error <- ggplot(prop_emerged_female, aes(x=temp, y=`mean(emerged)`, group=emerg_or_TFL, ymin=lower, ymax=upper)) +
                         geom_point(position=pd0, size=1.5, aes(colour=factor(emerg_or_TFL))) +
                        # geom_errorbar(position=pd0, size=0.5, width=.15, aes(colour=factor(emerg_or_TFL)))+ #add in for error bars
                         geom_line(position=pd0, aes(colour=factor(emerg_or_TFL)))+
                         scale_x_continuous('Pupal heat-shock temperature (°C)',breaks = round(seq(min(28), max(35), by = 1),1)) +
                         scale_y_continuous('Proportion')+
                         coord_cartesian(ylim=c(0, 1)) +
                         scale_color_discrete("",labels=c("Emergence", "Female fertility"))+
                         theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                               axis.line=element_line(colour='gray90'),
                               axis.text=element_text(colour='black'),
                               panel.grid.major = element_line("gray90"),
                               axis.title.y = element_text(vjust=1.5),
                               plot.title = element_text(face="bold"),
                               plot.background = element_rect(fill = "transparent",colour = NA))

TFL_F_emergence_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_TFL_F_emergence_error.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(TFL_F_emergence_error)
#dev.off() 

#### 5. Female fertility with error bars ####

pd<-position_dodge(1)

TFL_F_error<- ggplot(prop_F_larval1, aes(x=temp, y=`mean(emerged)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=0.5, colour="black") +
  geom_line(position=pd, size=0.5, colour="black")+
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature (°C)', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of surviving females fertile')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

TFL_F_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_TFL_F_error.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(TFL_F_error)
#dev.off() 


#### 6. Male and female fertility vs emergence ####

prop_F_larval1<-cbind((binom.confint(prop_F_larval1$`mean(F_larval1)`*prop_F_larval1$N, prop_F_larval1$N, conf.level = 0.95, methods = "logit")),prop_F_larval1)

colnames(prop_F_larval1)[8] <- "mean(emerged)"
prop_emerged_all_male_female <- rbind(prop_emerged, prop_M_larval, prop_F_larval1)
prop_emerged_all_male_female [,"emerg_male_female"]<- c(0,0,0,0,0,0,0,0,1,1,1,1,1,1,1,2,2,2,2,2,2,2)
prop_emerged_all_male_female$emerg_male_female<-as.factor(prop_emerged_all_male_female$emerg_male_female)


pd3 <- position_dodge(0)
TFL_emergence_m_f <- ggplot(prop_emerged_all_male_female, aes(x=temp, y=`mean(emerged)`, group=emerg_male_female, ymin=lower, ymax=upper)) +
                       geom_point(position=pd3, size=1.5, aes(colour=factor(emerg_male_female))) +
                       #geom_errorbar(position=pd3, size=0.5, width=.15, aes(colour=factor(emerg_male_female)))+ #use this for error bars
                       geom_line(position=pd3, aes(colour=factor(emerg_male_female)))+
                       scale_x_continuous('Pupal heat-shock temperature (°C)',breaks = round(seq(min(28), max(35), by = 1),1)) +
                       scale_y_continuous('Proportion')+
                       coord_cartesian(ylim=c(0, 1)) +
                       scale_color_discrete("",labels=c("Emergence", "Male fertility", "Female fertility"))+
                       theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                             axis.line=element_line(colour='gray90'),
                             axis.text=element_text(colour='black'),
                             panel.grid.major = element_line("gray90"),
                             axis.title.y = element_text(vjust=1.5),
                             plot.title = element_text(face="bold"),
                             plot.background = element_rect(fill = "transparent",colour = NA))

TFL_emergence_m_f

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_male_female_fertility.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(TFL_emergence_m_f)
#dev.off() 


#### 7. Female survival throughout TFL experiment ####

#Create subset with femlaes that have the full data
subF_survival<-subset(pupaeTFL, F_female_survived_laying3==1 |F_female_survived_laying3==0)
#subF_survival$F_female_survived_laying3<-as.numeric(subF_survival$F_female_survived_mating)

#Gather useful columns and assing values based on days
x<- gather(subF_survival, Days, Survival, emerged, survived_to_mating, F_female_survived_mating, F_female_survived_laying1, F_female_survived_laying2, F_female_survived_laying3)
x$Days<-sapply(x$Days,switch,'emerged'=0, 'survived_to_mating'=5,'F_female_survived_mating'=6,'F_female_survived_laying1'=8,'F_female_survived_laying2'=10, 'F_female_survived_laying3'=12)
x$temp<-as.factor(x$temp)

#Work out proportions alive at each time point by temperature
x$Survival<-as.numeric(x$Survival)
prop_F_survival <- summarise(group_by(x, temp, Days), mean(Survival),N=length(Survival))

#Normal ggplot of survival over time
F_survival<- ggplot(prop_F_survival, aes(x=Days, y=`mean(Survival)`, group=temp)) +
  geom_point(aes(colour=temp)) +
  geom_line(aes(colour=temp)) +
  coord_cartesian(ylim=c(0, 1), xlim=c(0,13)) +
  scale_x_continuous('Days after emergence', breaks= round(seq(min(0), max(13), by = 1),1)) +
  scale_y_continuous('Proportion of females alive', breaks=round(seq(min(0), max(1), by =0.1),1)) +
  labs(colour = "Temperature (°C)") +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA))

F_survival

#### 8. Female survival through to end of first laying ####

#Create subset with femlaes that have the full data
subF_survival_FL<-subset(pupaeTFL, F_female_survived_laying1==1 |F_female_survived_laying1==0)
#subF_survival$F_female_survived_laying3<-as.numeric(subF_survival$F_female_survived_mating)

#Gather useful columns and assing values based on days
survival_x<- gather(subF_survival_FL, Days, Survival, emerged, survived_to_mating, F_female_survived_mating, F_female_survived_laying1)
survival_x$Days<-sapply(survival_x$Days,switch,'emerged'=0, 'survived_to_mating'=5,'F_female_survived_mating'=6,'F_female_survived_laying1'=8)
survival_x$temp<-as.factor(survival_x$temp)

#Work out proportions alive at each time point by temperature
survival_x$Survival<-as.numeric(survival_x$Survival)
prop_F_survival_FL <- summarise(group_by(survival_x, temp, Days), mean(Survival),N=length(Survival))

#Normal ggplot of survival over time
F_survival_FL <-ggplot(prop_F_survival_FL, aes(x=Days, y=`mean(Survival)`, group=temp)) +
                geom_point(aes(colour=temp)) +
                geom_line(aes(colour=temp)) +
                coord_cartesian(ylim=c(0, 1), xlim=c(0,8)) +
                scale_x_continuous('Days after emergence', breaks= round(seq(min(0), max(8), by = 1),1)) +
                scale_y_continuous('Proportion of females alive', breaks=round(seq(min(0), max(1), by =0.1),1)) +
                labs(colour = "Temperature (°C)") +
                theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                      axis.line=element_line(colour='gray90'),
                      axis.text=element_text(colour='black'),
                      panel.grid.major = element_line("gray90"),
                      axis.title.y = element_text(vjust=1.5),
                      plot.title = element_text(face="bold"),
                      plot.background = element_rect(fill = "transparent",colour = NA))

F_survival_FL

#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_female_survival.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(F_survival_FL)
#dev.off() 


#### 9. Male survival ####
#Create subset with males that have the full data
subM_survival<-subset(pupaeTFL, M_male_survived_mating==1 |M_male_survived_mating==0)
#subM_survival$M_male_survived_mating<-as.numeric(subM_survival$M_male_survived_mating)

#Gather useful columns and assing values based on days
survival_M<- gather(subM_survival, Days, Survival, emerged, survived_to_mating, M_male_survived_mating)
survival_M$Days<-sapply(survival_M$Days,switch,'emerged'=0, 'survived_to_mating'=3,'M_male_survived_mating'=4)
survival_M$temp<-as.factor(survival_M$temp)

#Work out proportions alive at each time point by temperature
survival_M$Survival<-as.numeric(survival_M$Survival)
prop_M_survival <- summarise(group_by(survival_M, temp, Days), mean(Survival),N=length(Survival))

#Normal ggplot of survival over time
M_survival <-ggplot(prop_M_survival, aes(x=Days, y=`mean(Survival)`, group=temp)) +
             geom_point(aes(colour=temp)) +
             geom_line(aes(colour=temp)) +
             coord_cartesian(ylim=c(0, 1), xlim=c(0,8)) +
             scale_x_continuous('Days after emergence', breaks= round(seq(min(0), max(8), by = 1),1)) +
             scale_y_continuous('Proportion of males alive', breaks=round(seq(min(0), max(1), by =0.1),1)) +
             labs(colour = "Temperature (°C)") +
            theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                  axis.line=element_line(colour='gray90'),
                  axis.text=element_text(colour='black'),
                  panel.grid.major = element_line("gray90"),
                  axis.title.y = element_text(vjust=1.5),
                  plot.title = element_text(face="bold"),
                  plot.background = element_rect(fill = "transparent",colour = NA))

M_survival

#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_male_survival.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(M_survival)
#dev.off() 


#### 10. Pupation length vs fertility ####

prop_M_larval_pupation <- as.data.frame(summarise(group_by(subpupaeTFL, temp, pupation), mean(M_larval), N=length(M_larval)))
#prop_M_larval_pupation <- prop_M_larval_pupation[-c(11, 12),]
prop_M_larval_pupation$number_pupated<-prop_M_larval_pupation$`mean(M_larval)`*prop_M_larval_pupation$N
prop_M_larval_pupation$pupation<-as.factor(prop_M_larval_pupation$pupation)
prop_M_larval_pupation <- cbind(prop_M_larval_pupation,binom.confint(prop_M_larval_pupation$number_pupated, prop_M_larval_pupation$N, conf.level = 0.95, methods = "logit"))





TFL_M_pupation<- ggplot(prop_M_larval_pupation, aes(x=temp, y=`mean(M_larval)`, group=pupation, ymax=upper, ymin=lower)) + 
  geom_point(size=1.5, aes(colour=factor(pupation))) +
  geom_line(aes(colour=factor(pupation)))+
  geom_errorbar(aes(colour=factor(pupation)), size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature (°C)', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of surviving males fertile')+
  scale_color_discrete("Pupation length")+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA))

TFL_M_pupation

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_TFL_M_pupation.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(TFL_M_pupation)
#dev.off() 





#### 11. Male offspring numbers ####

sub_M_adults<- subset(pupaeTFL, M_adults !='NA')

male_offspring <- ggplot(sub_M_adults, aes(x=temp, y=M_adults, group=factor(temp)))+
                  geom_point(position = position_jitter(w = 0.1, h = 0))+
                  geom_boxplot(outlier.colour=NA, alpha=.5, width=.5) +
                  scale_x_continuous("Pupal heat shock temperature (°C)",breaks = round(seq(min(28), max(35), by = 1),1))+
                  scale_y_continuous("Number of adult offspring produced by males")+
                  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                        axis.line=element_line(colour='gray90'),
                        axis.text=element_text(colour='black'),
                        panel.grid.major = element_line("gray90"),
                        axis.title.y = element_text(vjust=1.5),
                        plot.title = element_text(face="bold"),
                        plot.background = element_rect(fill = "transparent",colour = NA))

male_offspring

#### 12. Female offspring numbers laying 1 ####

sub_F_adults<- subset(pupaeTFL, F_adults1 !='NA')

female_offspring <- ggplot(sub_F_adults, aes(x=temp, y=F_adults1, group=factor(temp)))+
                    geom_point(position = position_jitter(w = 0.1, h = 0))+
                    geom_boxplot(outlier.colour=NA, alpha=.5, width=.5) +
                    scale_x_continuous("Pupal heat shock temperature (°C)",breaks = round(seq(min(28), max(35), by = 1),1))+
                    scale_y_continuous("Number of adult offspring produced by females")+
                    theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                          axis.line=element_line(colour='gray90'),
                          axis.text=element_text(colour='black'),
                          panel.grid.major = element_line("gray90"),
                          axis.title.y = element_text(vjust=1.5),
                          plot.title = element_text(face="bold"),
                          plot.background = element_rect(fill = "transparent",colour = NA))

female_offspring

#### 13. Male vs female offspring numbers (per laying female) ####

sub_all_adults<- subset(pupaeTFL, F_adults1!='NA' | M_adults!='NA')
sub_all_adults$M_adults<-as.numeric(sub_all_adults$M_adults)
sub_all_adults$M_female_survived_laying<-as.numeric(sub_all_adults$M_female_survived_laying)
sub_all_adults[,"Offspring_per_female"]<-sub_all_adults$M_adults/sub_all_adults$M_female_survived_laying


#Create a new column that includes both male (/2) and female (as is) adult numbers
sub_all_adults<- sub_all_adults %>% 
  replace_na(list(Offspring_per_female = "", F_adults1 = "")) %>% 
  unite(all_adults, Offspring_per_female, F_adults1, sep = "")


#Plot


sub_all_adults$temp<-as.factor(sub_all_adults$temp)
#sub_all_adults$temp<-as.numeric(sub_all_adults$temp)
sub_all_adults$all_adults<-as.numeric(sub_all_adults$all_adults)

sub_all_adults$sex <- factor(sub_all_adults$sex,levels = c(0,1),
labels = c("Female", "Male"))

all_offspring <- ggplot(sub_all_adults, aes(x=temp, y=all_adults))+
                    geom_point(position = position_jitterdodge(jitter.width = .2), aes(colour=sex))+
                    geom_boxplot(outlier.colour=NA, alpha=.5, width=.5, aes(fill=sex), show.legend = FALSE) +
                    scale_x_discrete('Pupal heat-shock temperature (°C)') +
                    scale_y_continuous('Offspring number per laying female')+
                     guides(color=guide_legend("Sex"))+
                    theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                          axis.line=element_line(colour='gray90'),
                          axis.text=element_text(colour='black'),
                          panel.grid.major = element_line("gray90"),
                          axis.title.y = element_text(vjust=1.5),
                          plot.title = element_text(face="bold"),
                          plot.background = element_rect(fill = "transparent",colour = NA))

all_offspring

#### 14. Male vs female numbers with 0s removed ####
sub_all_adults_no_zero<- subset(pupaeTFL, F_adults1>0 | M_adults>0)
sub_all_adults_no_zero$M_adults<-as.numeric(sub_all_adults_no_zero$M_adults)
sub_all_adults_no_zero$M_female_survived_laying<-as.numeric(sub_all_adults_no_zero$M_female_survived_laying)
sub_all_adults_no_zero[,"Offspring_per_female"]<-sub_all_adults_no_zero$M_adults/sub_all_adults_no_zero$M_female_survived_laying


#Create a new column that includes both male (/2) and female (as is) adult numbers
sub_all_adults_no_zero<- sub_all_adults_no_zero %>% 
  replace_na(list(Offspring_per_female = "", F_adults1 = "")) %>% 
  unite(all_adults, Offspring_per_female, F_adults1, sep = "")


#Plot


sub_all_adults_no_zero$temp<-as.factor(sub_all_adults_no_zero$temp)
#sub_all_adults_no_zero$temp<-as.numeric(sub_all_adults_no_zero$temp)
sub_all_adults_no_zero$all_adults<-as.numeric(sub_all_adults_no_zero$all_adults)

sub_all_adults_no_zero$sex <- factor(sub_all_adults_no_zero$sex,levels = c(0,1),
labels = c("Female", "Male"))

all_offspring_no_zero <- ggplot(sub_all_adults_no_zero, aes(x=temp, y=all_adults))+
                    geom_point(position = position_jitterdodge(jitter.width = .2), aes(colour=sex))+
                    geom_boxplot(outlier.colour=NA, alpha=.5, width=.5, aes(fill=sex), show.legend = FALSE) +
                    scale_x_discrete('Pupal heat-shock temperature (°C)') +
                    scale_y_continuous('Offspring number per laying female')+
                     guides(color=guide_legend("Sex"))+
                    theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                          axis.line=element_line(colour='gray90'),
                          axis.text=element_text(colour='black'),
                          panel.grid.major = element_line("gray90"),
                          axis.title.y = element_text(vjust=1.5),
                          plot.title = element_text(face="bold"),
                          plot.background = element_rect(fill = "transparent",colour = NA))

all_offspring_no_zero


#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D-mel_all_offspring.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(all_offspring)
#dev.off() 


#### 15. Male and female offspring temp as numeric ####
sub_all_adults$temp<-as.character(sub_all_adults$temp)
sub_all_adults$temp<-as.numeric(sub_all_adults$temp)

all_offspring_temp_numeric <- ggplot(sub_all_adults, aes(x=temp, y=all_adults, group=sex))+
                          stat_summary(fun.data= mean_se, geom="pointrange", aes(colour=sex))+
                          stat_summary(fun.y = mean, geom="line", aes(colour=sex))+
                          scale_x_continuous('Pupal heat shock temperature (°C)') +
                          scale_y_continuous('Offspring number per laying female') +
                          scale_color_discrete("Sex") +
                          theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                                axis.line=element_line(colour='gray90'),
                                axis.text=element_text(colour='black'),
                                panel.grid.major = element_line("gray90"),
                                axis.title.y = element_text(vjust=1.5),
                                plot.title = element_text(face="bold"),
                                plot.background = element_rect(fill = "transparent",colour = NA))
all_offspring_temp_numeric

#png(filename="C:/Users/bwalsh/Documents/R docs/D-melall_offspring_temp_numeric.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(all_offspring_temp_numeric)
#dev.off() 

# EXPERIMENT 2: BEHAV. & MOTILITY ####

# Reading data in and subsets ####

beh.mot<-read.csv("D.mel_behav_motility_data.csv", header=TRUE)
#beh.mot$motility<- as.factor(beh.mot$motility)

#Subsets
submate<- subset(beh.mot,mating_success ==1 |mating_success ==0)
submatefemale<- subset(submate, sex==0)
submatemale<- subset(submate, sex==1)
submotility_3<- subset(beh.mot,motility_3==1 |motility_3==0)
submotility_8<- subset(beh.mot,motility_8==1 |motility_8==0)
submotility_14<-subset(beh.mot,motility_14==1|motility_14==0)

#### Calculate proportions####

# Male matings
prop_male_mating <- summarise(group_by(submatemale, temp), mean(mating_success), N=length(mating_success))
prop_male_mating<-cbind((binom.confint(prop_male_mating$`mean(mating_success)`*prop_male_mating$N, prop_male_mating$N, conf.level = 0.95, methods = "logit")),prop_male_mating)

# Female matings

prop_female_mating <- summarise(group_by(submatefemale, temp), mean(mating_success), N=length(mating_success))
prop_female_mating<-cbind((binom.confint(prop_female_mating$`mean(mating_success)`*prop_female_mating$N, prop_female_mating$N, conf.level = 0.95, methods = "logit")),prop_female_mating)

#Calculate proportions of 3 day motility with error

prop_motile_3 <- summarise(group_by(submotility_3, temp), mean(motility_3), N=length(motility_3))
prop_motile_3 <-cbind((binom.confint(prop_motile_3$`mean(motility_3)`*prop_motile_3$N, prop_motile_3$N, conf.level = 0.95, methods = "logit")),prop_motile_3)


prop_motile_8 <- summarise(group_by(submotility_8, temp), mean(motility_8), N=length(motility_8))
prop_motile_8 <-cbind((binom.confint(prop_motile_8$`mean(motility_8)`*prop_motile_8$N, prop_motile_8$N, conf.level = 0.95, methods = "logit")),prop_motile_8)

prop_motile_14 <- summarise(group_by(submotility_14, temp), mean(motility_14), N=length(motility_14))
prop_motile_14 <-cbind((binom.confint(prop_motile_14$`mean(motility_14)`*prop_motile_14$N, prop_motile_14$N, conf.level = 0.95, methods = "logit")),prop_motile_14)



# PLOTS ####
#### 1. Male mating with error ####
pd<-position_dodge(1)

male_mating_error<- ggplot(prop_male_mating, aes(x=temp, y=`mean(mating_success)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=1.5, colour="black") +
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of treated males successfully mating')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

male_mating_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D.mel male mating.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(male_mating_error)
#dev.off() 



#### 2. Female mating with error ####
pd<-position_dodge(1)

female_mating_error<- ggplot(prop_female_mating, aes(x=temp, y=`mean(mating_success)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=1.5, colour="black") +
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of treated females successfully mating')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

female_mating_error

#Saving high res plot
#png(filename="C:/Users/bwalsh/Documents/R docs/D.mel female mating.png", type="cairo", units="in", width=8, height=6, pointsize=10, res=1000)
#print(female_mating_error)
#dev.off() 


#### 3. 3 day motility with error ####

pd<-position_dodge(1)

motility_3_error<- ggplot(prop_motile_3, aes(x=temp, y=`mean(motility_3)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=1.5, colour="black") +
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of males with motile sperm')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

#motility_3_error

#### 4. 8 day motility with error ####

motility_8_error<- ggplot(prop_motile_8, aes(x=temp, y=`mean(motility_8)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=1.5, colour="black") +
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of males with motile sperm')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

#motility_8_error

#### 5. All motility with error ####
colnames(prop_motile_8)[8] <- "mean(motility_3)"
colnames(prop_motile_14)[8] <- "mean(motility_3)"
prop_motile_all <- rbind(prop_motile_3,prop_motile_8,prop_motile_14)
prop_motile_all ["days"]<- c(3,3,8,8,14,14)
prop_motile_all$days<-as.factor(prop_motile_all$days)



pd1 <- position_dodge(0.25)
motility_all_error <- ggplot(prop_motile_all, aes(x=days, y=`mean(motility_3)`, group=temp, ymin=lower, ymax=upper)) + 
  geom_point(position=pd1, size=1.5, aes(colour=factor(temp))) +
  geom_errorbar(position=pd1, size=0.5, width=.15, aes(colour=factor(temp)))+
  geom_line(position=pd1, aes(colour=factor(temp)))+
  scale_x_discrete('Days after emergence') +
  scale_y_continuous('Proportion of males with motile sperm')+
  scale_color_manual('Temperature',values=c("#00BFC4", "#F8766D"))+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA))

motility_all_error



#### 6. Motility of block 5 IN pupaeTFL dataset ####
#These flies have mated, then 24 hours to recover, then tested for motility

submot_TFL<- subset(pupaeTFL,M_motility==1 |M_motility==0)

prop_mot_TFL <- summarise(group_by(submot_TFL, temp), mean(M_motility), N=length(M_motility))
prop_mot_TFL <-cbind((binom.confint(prop_mot_TFL$`mean(M_motility)`*prop_mot_TFL$N, prop_mot_TFL$N, conf.level = 0.95, methods = "logit")),prop_mot_TFL)

pd<-position_dodge(1)

motility_TFL_error<- ggplot(prop_mot_TFL, aes(x=temp, y=`mean(M_motility)`, ymin=lower, ymax=upper)) + 
  geom_point(position=pd, size=1.5, colour="black") +
  geom_errorbar(colour="black", position=pd, size=0.5, width=.2)+
  scale_x_continuous('Pupal heat shock temperature', breaks= round(seq(min(28), max(35), by = 1),1)) +
  scale_y_continuous('Proportion of males with motile sperm')+
  coord_cartesian(ylim=c(0, 1)) +
  theme(panel.background = element_rect(fill = "white",colour = "gray90"),
        axis.line=element_line(colour='gray90'),
        axis.text=element_text(colour='black'),
        panel.grid.major = element_line("gray90"),
        axis.title.y = element_text(vjust=1.5),
        plot.title = element_text(face="bold"),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position='none')

motility_TFL_error

#### 7. Female mating latency temp as a numeric ####

submatefemale$temp<- as.numeric(submatefemale$temp)
mating_latency_F<- ggplot(submatefemale, aes(x=temp, y=latency_s))+
                          stat_summary(fun.data= mean_se, geom="pointrange")+
                          scale_x_continuous('Pupal heat shock temperature (°C)') +
                          scale_y_continuous('Female mating latency (s)') +
                          theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                                axis.line=element_line(colour='gray90'),
                                axis.text=element_text(colour='black'),
                                panel.grid.major = element_line("gray90"),
                                axis.title.y = element_text(vjust=1.5),
                                plot.title = element_text(face="bold"),
                                plot.background = element_rect(fill = "transparent",colour = NA),
                                legend.position='none')
mating_latency_F

#### 8. Female mating latency temp as a factor ####



submatefemale$temp<-as.factor(submatefemale$temp)
#sub_all_adults$temp<-as.numeric(sub_all_adults$temp)
submatefemale$latency_s<-as.numeric(submatefemale$latency_s)

#sub_all_adults$sex <- factor(sub_all_adults$sex,levels = c(0,1),
#labels = c("Female", "Male"))

mating_latency_F_factor <- ggplot(submatefemale, aes(x=temp, y=latency_s))+
                          geom_point(position = position_jitter(w=.1))+
                          geom_boxplot(outlier.colour=NA, alpha=.5, width=.2, aes(fill=sex), show.legend = FALSE) +
                          scale_x_discrete('Pupal heat-shock temperature of females (°C)') +
                          scale_y_continuous('Mating latency (s)', breaks= round(seq(min(0), max(9000), by = 1000),1))+
                          #guides(color=guide_legend("Sex"))+
                          theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                                axis.line=element_line(colour='gray90'),
                                axis.text=element_text(colour='black'),
                                panel.grid.major = element_line("gray90"),
                                axis.title.y = element_text(vjust=1.5),
                                plot.title = element_text(face="bold"),
                                plot.background = element_rect(fill = "transparent",colour = NA))
  
mating_latency_F_factor


#### 9. Female mating duration temp as a factor ####

submatefemale$temp<-as.factor(submatefemale$temp)
#sub_all_adults$temp<-as.numeric(sub_all_adults$temp)
submatefemale$duration_s<-as.numeric(submatefemale$duration_s)

#sub_all_adults$sex <- factor(sub_all_adults$sex,levels = c(0,1),
#labels = c("Female", "Male"))

mating_duration_F_factor <- ggplot(submatefemale, aes(x=temp, y=duration_s))+
                          geom_point(position = position_jitter(w=.1))+
                          geom_boxplot(outlier.colour=NA, alpha=.5, width=.2, aes(fill=sex), show.legend = FALSE) +
                          scale_x_discrete('Pupal heat-shock temperature of females (°C)') +
                          scale_y_continuous('Mating duration (s)', breaks= round(seq(min(0), max(2000), by = 200),1))+
                          #guides(color=guide_legend("Sex"))+
                          theme(panel.background = element_rect(fill = "white",colour = "gray90"),
                                axis.line=element_line(colour='gray90'),
                                axis.text=element_text(colour='black'),
                                panel.grid.major = element_line("gray90"),
                                axis.title.y = element_text(vjust=1.5),
                                plot.title = element_text(face="bold"),
                                plot.background = element_rect(fill = "transparent",colour = NA))
  
mating_duration_F_factor




```
This document is a summary of the experimental work I have completed as of `r format(Sys.time(), '%d %B %Y')`, including a brief description of the methodology used, plots of the data collected, and future directions in which the work could be taken.

The experiment was designed to examine the Thermal Fertility Limits (TFL) of Drosophila melanogaster, when exposed to a heat shock during the pupal stage.

# 1. Methods
We subjected *Drosophila melanogaster* pupae to a 48 hour heat-shock at static temperatures ranging from 28&deg;C to 35&deg;C. Individuals that had pupated within 8 hours of each other were transferred to 96 well PCR plates (around 4-8 pupae per well), and subjected to one of 8 temperatures. For temperatures below 34&deg;C, gradients creating distinct temperatures at specific locations across plates were used (gradient 1: 28,30,32.5&deg;C; gradient 2:29,31,33.5&deg;C).In each of these plates, intermediate temperatures were not used, giving 3 distinct temperatures in each case. For temperatures 34&deg;C and 35&deg;C, full static temperature plates were used in order to maximise sample size due to low survival. Plates were covered with standard PCR plate seals, with holes pierced in each well to allow for gaseous exchange. After 48 hours at their respective temperatures, the pupae were transferred to vials according to the PCR column they were treated in, placed back at 25&deg;C and left to emerge.

Vials were checked 3 times per day, and both males and females were separated at emergence and kept at 25&deg;C. Thereafter, the project was split into two distinct experiments.

### Male fertility
Males were kept for 3 days until sexual maturity, and subsequently placed with 2 females for 24 hours to allow males to mate. After this, males were discarded and females given 3 days to lay eggs. After laying, females were discarded and the vials were monitored for the presence of larvae (0/1), and the number of emerging adults in each vial was counted.

### Female fertility
Females were given 5 days to reach sexual maturity, and then given 2 males for 24 hours to allow females an opportunity to mate. After this, males were discaded and females given 3 opportunities to lay, each 48 hours to allow for accurate egg counts. This gave 3 egg counts, larval scores, and adult counts per laying attempt. It should be noted here that female survival was also monitored at each stage of this process, giving 6 survival data points per individual.

# 2. Results
Emergence of individuals from the treatments is shown below (Fig. 1). There was a relatively high emergence from 28&deg;C to 32.5&deg;C, and after this emergence drops significantly to 0 at 35&deg;C. Due to the difficulty of identifying the sex of an individual at the pupal stage, this data cannot be split into male/ female emergence.

```{r,echo=FALSE,fig.cap="**Figure 1** The probability of emergence of *D. melanogaster* individuals exposed to a 48 hour heat shock at the pupal stage. Error bars represent 95% confidence intervals."}
plot(emerged_error)
```

### Male fertility
Male fertility followed a similar trend to emergence with a relatively steep decline after a high plateau, however this decline occurs at much lower temperatures when examining fertilty as a binary trait. Figure 2 shows the proportion of fertile male *D. melanogaster* individuals over the range of pupal heat-shock temperatures. This indicates that a male's fertility is almost fully compromised at 32.5&deg;C, a temperature with around 75% emergence. 

```{r,echo=FALSE,fig.cap="**Figure 2** The proportion of fertile male *D. melanogaster* individuals that have been exposed to a 48 hour heat shock at the pupal stage. Error bars represent 95% confidence intervals."}
plot(TFL_error)
```

Figure 3 combines the two above graphs to show this difference in these two traits. There is around a 2&deg;C difference in where these two traits are below 20%. This gap between fertility and mortality is important, and suggests that using mortality limits may underestimate the impact of pupal temperature on individuals in this species.

```{r, echo = FALSE, fig.cap="**Figure 3** The proportion of two traits of *D. melanogaster* that have been exposed to a 48 hour heat shock during the pupal stage. The red line represents the proportion of flies that emerge, and the green line represents the fertility of emerging **males**."}
plot(TFL_emergence_error)
```



### Female fertility
The data for female fertility from the larval measure is much less clear, partly due to the fact that I have had to tinker with the methods to get the experiment to work properly and as such the data is not as robust. However, it appears that there is a slight impairment of fertility as temperature increases, as shown in the graph below.

```{r, echo=FALSE, fig.cap="**Figure 4** The proportion of two traits of *D. melanogaster* that have been exposed to a 48 hour heat shock during the pupal stage. The red line represents the proportion of flies that emerge, and the green line represents the fertility of emerging **females**."}
plot(TFL_emergence_m_f)
```

### Female survival
Of the flies that have successfully emerged, it appears that the majority of individuals, regardless of treatment, can survive to sexual maturity (which we have allocated as 5 days in *D. melanogaster*). After mating, the females are given 48 hours to lay 3 separate times. This gave us 6 data points for survival per individual, which allowed me to examine survival of females throughout the experiment. Figure 5 shows that immediately after mating and during the first laying period (between days 6 and 8), a significant proportion of females exposed to higher pupal temperatures die, suggesting that there could be an extreme cost of mating when an indvidual is already heat-stressed from an earlier life-history stage.

This would have to be tested with a much more thorough experimental design for the question to be answered effectively. Also, it should be noted that although the sample size for the temperatures 28&deg;C, 30&deg;C, and 32.5&deg;C are 21 each, the sample size for 34?C at this point is only 4 due to the low emergence of individuals at this temperature.

```{r, echo=FALSE, fig.cap= "**Figure 5** Percentage survival of female *D. melanogaster* individuals that have been exposed to a heat shock at the pupal stage. Females were mated between days 5 and 6."}
plot(F_survival)
```

# 3. Following D. melanogaster experiments

### Finishing this experiment

In my opinion, the *D. melanogaster* pupal TFL experiment is close to being wrapped up, with perhaps 1 or 2 more blocks to go if we decide to add more temperatures. Including more temperatures might be a good idea, so we can more precicely predict where emergence and fertility is dropping off. The temperatures we could add would be 29, 31, and 33. This could allow us to see how sharp the fall off in fertility is, and will give us fertility measures for temperatures that exhibit varying levels of stress (i.e. 50% fertility at X&deg;C vs 25% fertility at X&deg;C), which we can use in future experiments (see below).

### Experiment 2: Mating behaviour & sperm motility

I am currently running this, and should have the data by 30/4/18. At this point, we know that pupal heat shocks of temperatures of 32.5&deg;C render around 80% of individuals sterile. What we do not know is where this sterility is coming from. This experiment subjects *D. melanogaster* larvae to two temperatures: 28&deg;C (benign) and 32.5&deg;C (stressful) as above, and I will be assessing:

*i). Behaviour*
Simply a mating assay: do the flies mate? All of the usual traits will be measured (success, latency, duration).

*ii). Sperm motility*
The technique that the French group (led by Jean David, Dominique Joly) used was to dissect the male's seminal vesicle and assess whether there is *any* sign of motility (as a 1/0 binary trait). I have been able to confidently do this in the lab on benign individuals, seeing motility on all males examined so far. Thus, any noticible lack of motility should be a useful indicator of sperm viability as a qualitative trait. This is an easy trait to measure, however if motility is not shown to be a truly binary trait (which is likely), then there may be problems with this technique. David et al do not attempt to quantify any quantitative differences in sperm motility-it may be useful to attempt to design a qualitative or quantitative scale for motility if this technique (1/0) is not useful.

### Experiment 3: Female survival & mating stress

Following on from the survival data in Figure 5, another possible follow-on is to examine how female survival changes with temperature, and more interestingly, does mating induce a significant cost to stressed flies and result in death, as our data suggest?
### Experiment 4: Recovery of male sterility
It has been shown that *D. melanogaster* recover some fertilty over time after heat stress (Chakir et al 2002). If given new females every day for over a week after a pupal heat-stress, we could examine whether this reduction in fertility is permanent or only temporary, which would give us an idea as to how severe it could be for populations.

# 4. Future multi-species experiments

The most obvious major experiment to attempt when I have examined *D. melanogaster* in some depth is to take this method and apply it to a large number of species that spans the phylogenetic range of Drosophilidae, in order to examine whether these traits are variable and/or phylogenetically constrained, as well as to examine whether they correlate with species distributions or any other aspects of their biology/ecology.

Problems with applying this method to different species will be many, especially as I am treating them at a particular stage of development. Most obviously, the different species develop and mature at different rates, so the time between heat-shock and assessing fertility will vary from species to species.

_Species we currently have in the lab_

|Melanogaster|Obscura|Virilis|Repleta|
|---|---|---|---|
|D. melanogaster|D. subobscura|D. virilis|D. repleta|
|D. simulans (Madeira)|D. pseudoobascura|D. novamexicana|D. hydei|
|D. suzukii|D. bifasciatus|D. flavomontana|D. mercatorum|
|D. ananassae|D. maderensis|||